---
title: "Clinical Study Protocol"
subtitle: "Adapted to illustrate agentic code review with Claude Code"
author: "Example Author"
date: today
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(gt)

# Reusable gt style helper (mirrors HARPER template)
gt_style <- function(x) {
  gt::tab_options(x, data_row.padding = gt::px(3))
}
```

> **Important:** This is a **mock protocol** adapted from a published clinical study
> protocol solely to illustrate agentic code review and validation with
> [Claude Code](https://claude.ai/code). Drug names and identifying information
> have been replaced with placeholders (**Drug A**, **Drug B**). All data are
> fully simulated. No real-world clinical conclusions should be drawn.

This protocol is based on the *HARmonized Protocol Template to Enhance
Reproducibility (HARPER) of Hypothesis Evaluating Real-World Evidence Studies on
Treatment Effects: A Good Practices Report of a Joint ISPE/ISPOR Task Force*
(**HARPER**) protocol by Wang SV et al.

{{< pagebreak >}}

# Title page {.unnumbered}

## Title {.unnumbered .unlisted}

\[Mock Study\] Comparative Effectiveness of Drug A vs. Drug B — Adapted to
Illustrate Agentic Code Review

## Research question and objectives {.unnumbered .unlisted}

To evaluate the comparative effect of **Drug A vs. Drug B** on the composite
primary endpoint of all-cause mortality or hospitalization for heart failure in a
simulated cohort.

## Protocol version {.unnumbered .unlisted}

Version 1

## Contributors {.unnumbered .unlisted}

**Primary investigator contact information:** \[REDACTED\]

**Contributor names:** \[REDACTED\]

## Study registration {.unnumbered .unlisted}

**Site:** N/A

**Identifier:** N/A

## Sponsor {.unnumbered .unlisted}

**Organization:** N/A

**Contact:** N/A

## Conflict of interest {.unnumbered .unlisted}

\[REDACTED\]

{{< pagebreak >}}

# Abstract

This is a non-randomized, non-interventional **mock study** designed to
illustrate a complete target trial emulation workflow using fully simulated data.
The study design is adapted from a published new-user active-comparator cohort
study comparing Drug A versus Drug B on the composite end point of all-cause
mortality or hospitalization for heart failure.

Drug A (treatment = 1) and Drug B (treatment = 0) are used as generic
placeholders. The simulated cohort (n = 28,100) deliberately introduces
**confounding by indication** so that propensity-score overlap weighting has a
meaningful role to play. The primary analysis uses the **average treatment effect
in the overlap (ATO) population**, estimated via a weighted Cox
proportional-hazards model.

# Amendments and updates

No amendments have been made to this mock protocol.

```{r}
#| label: tbl-amendments
#| tbl-cap: "Amendments and updates"

dplyr::tibble(
  `Version date`    = character(0),
  `Version number`  = character(0),
  `Section`         = character(0),
  `Amendment`       = character(0),
  `Reason`          = character(0)
) |>
  gt::gt() |>
  gt_style()
```

# Milestones

```{r}
#| label: tbl-milestones
#| tbl-cap: "Milestones"

dplyr::tibble(
  Milestone = c(
    "Protocol finalisation",
    "Data simulation and cohort generation",
    "Primary endpoint analysis",
    "Report/HTML rendered"
  ),
  Date = c(
    "2025-02-14",
    "2026-06-14",
    "2026-06-14",
    "2026-06-14"
  )
) |>
  gt::gt() |>
  gt_style()
```

# Rationale and background

**What is known about the condition:** Heart failure with preserved ejection
fraction (HFpEF) carries high morbidity and mortality; patients with
concomitant type 2 diabetes mellitus (T2DM) face an especially elevated risk
of hospitalization and death.

**What is known about the exposure of interest:** Drug A is a newer dual-receptor
agonist; Drug B is an established single-receptor agonist. Both are approved for
glucose control in T2DM. Clinical guidelines consider them interchangeable for
the relevant indication during the study period, making an active-comparator
design appropriate.

**Gaps in knowledge:** Whether Drug A confers a differential benefit over Drug B
on cardiovascular outcomes — specifically HF hospitalization or all-cause
mortality — has not been established in randomized trials powered for this
composite endpoint.

**What is the expected contribution of this study?** This mock emulation
illustrates how a new-user active-comparator design with propensity-score overlap
weighting can address confounding by indication in a real-world evidence setting.
It is intended as a teaching example for agentic code review, not to generate
clinical evidence.

# Research question and objectives

```{r}
#| label: tbl-objectives
#| tbl-cap: "Research questions and objectives"
#| tbl-subcap:
#|   - "A. Primary research question and objective"
#|   - "B. Secondary research question and objective"
#| layout-ncol: 1

# A. Primary
dplyr::tibble(
  Element = c(
    "Objective",
    "Hypothesis",
    "Population",
    "Exposure",
    "Comparator",
    "Outcome",
    "Time",
    "Setting",
    "Main measure of effect"
  ),
  Definition = c(
    "To evaluate the comparative effect of Drug A vs. Drug B on all-cause mortality or heart failure hospitalization.",
    "Drug A will not meaningfully differ from Drug B in preventing the primary composite endpoint.",
    "Simulated new users of Drug A or Drug B (n = 28,100); age 30–90 years; T2DM setting.",
    "Drug A (treatment = 1).",
    "Drug B (treatment = 0).",
    "Composite: all-cause mortality OR hospitalization for heart failure (whichever occurs first).",
    "From 1 day after cohort entry until: composite event, informative censoring, or day 365.",
    "Inpatient (HF hospitalization) and all-cause mortality.",
    "Hazard ratio (HR) with 95% confidence interval."
  )
) |>
  gt::gt() |>
  gt_style()

# B. Secondary (not implemented in mock study)
dplyr::tibble(
  Element = c("Note"),
  Definition = c(
    "Secondary endpoints are not implemented in this mock study. See mock-study/03_primary_endpoint_analysis.qmd."
  )
) |>
  gt::gt() |>
  gt_style()
```

# Research methods

## Study design

**Research design:** New-user active-comparator cohort study.

**Rationale for study design choice:** The new-user active-comparator design
addresses confounding by indication through judicious selection of a comparator
group and subsequent statistical balancing of measured confounders via a
propensity score. Restricting to new users avoids prevalent-user bias (depletion
of susceptibles). The overlap (ATO) estimand is chosen because both treatment
groups are represented throughout the propensity-score distribution, making
equipoise-based comparison most meaningful.

## Study design diagram

The study window spans from a 365-day washout period before index date (time 0)
through up to 365 days of follow-up. Baseline covariates are assessed in the
window \[–365, 0\]; treatment is assigned at time 0 (first dispensing); the
primary composite event or censoring occurs during \[1, 365\].

## Setting

### Context and rationale for definition of time 0

Time 0 is defined as the date of first dispensing of Drug A or Drug B for each
patient. The new-user requirement is operationalized by a 12-month washout window
with no prior dispensing of either drug.

```{r}
#| label: tbl-entry
#| tbl-cap: "Operational Definition of Time 0 (index date) and other primary time anchors"

dplyr::tibble(
  `Study population`     = c("Drug A initiators", "Drug B initiators"),
  `Time anchor`          = c("First dispensing of Drug A", "First dispensing of Drug B"),
  `Number of entries`    = c("Single entry", "Single entry"),
  `Type of entry`        = c("Incident use", "Incident use"),
  `Washout window (days)`= c("[-365, -1]", "[-365, -1]"),
  `Variable in data`     = c("`treatment = 1`", "`treatment = 0`")
) |>
  gt::gt() |>
  gt_style()
```

### Context and rationale for study inclusion criteria

Patients are required to be new users (no prior dispensing of Drug A or Drug B
within 365 days before index date), aged 30–90 years, and to have a documented
chronic disease (T2DM setting). These criteria are simulated in
`generate_cohort()` via the age clipping and treatment-assignment mechanism.

```{r}
#| label: tbl-inclusion
#| tbl-cap: "Operational Definitions of Inclusion Criteria"

dplyr::tibble(
  Criterion          = c("New user", "Age"),
  Details            = c(
    "No dispensing of Drug A or Drug B in the 365-day washout window",
    "Age 30–90 years at cohort entry (simulated via `rnorm(n, 65, 10)` clipped to [30, 90])"
  ),
  `Order of application` = c("Before index date", "At index date"),
  `Assessment window`    = c("[-365, -1]", "[0, 0]"),
  `Applied to`           = c("Both arms", "Both arms")
) |>
  gt::gt() |>
  gt_style()
```

### Context and rationale for study exclusion criteria

For this mock study, no explicit exclusion criteria are implemented beyond the
new-user and age inclusion criteria above. In the source protocol, patients were
excluded for prior use of the comparator drug, end-stage organ disease, and
certain contraindications.

```{r}
#| label: tbl-exclusion
#| tbl-cap: "Operational Definitions of Exclusion Criteria"

dplyr::tibble(
  Criterion = c("(None implemented in mock)"),
  Details   = c("The mock study omits disease-specific exclusion criteria present in the source protocol."),
  `Assessment window` = c("N/A"),
  `Applied to`        = c("N/A")
) |>
  gt::gt() |>
  gt_style()
```

## Variables

### Context and rationale for exposure(s) of interest

Treatment group is assigned at time 0 via a logistic-regression-based
propensity mechanism that embeds confounding by indication: Drug A (the newer
agent) is preferentially initiated in younger patients with higher BMI and lower
comorbidity burden. This mimics real-world prescribing behaviour and makes
PS-based confounding adjustment necessary.

**Algorithm to define duration of exposure effect:** A 45-day grace period is
applied. Informative censoring is modelled as exponential dropout with mean 400
days, representing disenrollment or treatment discontinuation.

```{r}
#| label: tbl-exposure
#| tbl-cap: "Operational Definitions of Exposure"

dplyr::tibble(
  `Exposure group`   = c("Drug A", "Drug B"),
  `R variable`       = c("`treatment = 1`; `treatment_label = 'Drug A'`",
                          "`treatment = 0`; `treatment_label = 'Drug B'`"),
  `Type of entry`    = c("Incident (new user)", "Incident (new user)"),
  `Washout window`   = c("[-365, -1]", "[-365, -1]"),
  `Grace period`     = c("45 days", "45 days")
) |>
  gt::gt() |>
  gt_style()
```

### Context and rationale for outcome(s) of interest

The primary outcome is the **composite of all-cause mortality or hospitalization
for heart failure**, whichever occurs first. This is simulated using a Weibull
proportional-hazards model (`simsurv::simsurv`) with a baseline scale calibrated
to produce a 1-year cumulative incidence of approximately 3.4% in the Drug B arm,
matching the published source study.

```{r}
#| label: tbl-outcome
#| tbl-cap: "Operational Definitions of Outcome"

dplyr::tibble(
  `Outcome`          = c("HF hospitalization or all-cause mortality (composite)"),
  `Primary?`         = c("Yes"),
  `Type`             = c("Time-to-event"),
  `R variable`       = c("`status` (1 = event, 0 = censored); `time` (follow-up days)"),
  `Washout window`   = c("N/A"),
  `Simulation model` = c("Weibull(lambda = 1.8e-5, gamma = 1.15); maxt = 365 days")
) |>
  gt::gt() |>
  gt_style()
```

### Context and rationale for follow up

Follow-up begins 1 day after cohort entry (time 0) and ends at the earliest of:
the primary composite event, informative censoring (dropout), or the
administrative end of study at day 365. A 45-day grace period is allowed for
treatment gaps.

```{r}
#| label: tbl-follow_up
#| tbl-cap: "Operational Definitions of Follow up"

dplyr::tibble(
  `Rule`                           = c(
    "Follow-up start",
    "Primary composite event",
    "Informative censoring (dropout)",
    "Administrative censoring"
  ),
  `Definition`                     = c(
    "Day 1 after cohort entry (time 0 + 1 day)",
    "First occurrence of HF hospitalization or all-cause mortality",
    "Exponential(rate = 1/400); models disenrollment or discontinuation within the 45-day grace period",
    "Day 365 (52 weeks)"
  ),
  `R implementation`               = c(
    "`time = min(event_time, cens_time, 365)`",
    "`status = 1` if event_time ≤ cens_time AND ≤ 365",
    "`cens_time ~ rexp(n, rate = 1/400)`",
    "`maxt = 365` in `simsurv::simsurv()`"
  )
) |>
  gt::gt() |>
  gt_style()
```

### Context and rationale for covariates

All covariates are measured at cohort entry (baseline) and are used in both the
propensity-score logistic regression model and the Table 1 balance assessment.
Covariates were selected to represent the confounder domains adjusted for in the
source study: demographics, cardiometabolic burden, diabetes complications,
cardiovascular/kidney disease, other comorbidities, concomitant medications, and
healthcare utilization.

```{r}
#| label: tbl-covariates
#| tbl-cap: "Operational Definitions of Covariates"

dplyr::tribble(
  ~Domain,                     ~Variable,              ~Label,                             ~Type,       ~Distribution,
  "Demographics",              "`age`",                "Age, years",                       "Continuous","Normal(65, 10), clipped [30, 90]",
  "Demographics",              "`female`",             "Female sex",                       "Binary",    "Bernoulli(0.48)",
  "Demographics",              "`race`",               "Race",                             "Categorical","White 70%, Black 18%, Other 12%",
  "Demographics",              "`region`",             "US region",                        "Categorical","NE 20%, MW 23%, S 38%, W 19%",
  "Cardiometabolic burden",    "`bmi`",                "Body mass index, kg/m²",           "Continuous","Normal(33, 5), clipped [18, 60]",
  "Cardiometabolic burden",    "`prior_hf_hosp`",      "Prior HF hospitalization",         "Binary",    "Bernoulli(0.06)",
  "Cardiometabolic burden",    "`diuretic_use`",       "Diuretic use",                     "Binary",    "Bernoulli(0.28)",
  "Diabetes complications",    "`nephropathy`",        "Diabetic nephropathy",             "Binary",    "Bernoulli(0.15)",
  "Diabetes complications",    "`neuropathy`",         "Diabetic neuropathy",              "Binary",    "Bernoulli(0.22)",
  "Diabetes complications",    "`retinopathy`",        "Diabetic retinopathy",             "Binary",    "Bernoulli(0.12)",
  "Diabetes complications",    "`foot_ulcer`",         "Foot ulcer",                       "Binary",    "Bernoulli(0.05)",
  "Diabetes complications",    "`insulin_use`",        "Insulin use",                      "Binary",    "Bernoulli(0.30)",
  "CV / kidney disease",       "`pad`",                "Peripheral arterial disease",      "Binary",    "Bernoulli(0.10)",
  "CV / kidney disease",       "`afib`",               "Atrial fibrillation",              "Binary",    "Bernoulli(0.12)",
  "CV / kidney disease",       "`cardiomyopathy`",     "Cardiomyopathy",                   "Binary",    "Bernoulli(0.07)",
  "CV / kidney disease",       "`ckd`",                "Chronic kidney disease",           "Binary",    "Bernoulli(0.20)",
  "Other comorbidities",       "`copd`",               "COPD",                             "Binary",    "Bernoulli(0.13)",
  "Other comorbidities",       "`depression`",         "Depression",                       "Binary",    "Bernoulli(0.18)",
  "Other comorbidities",       "`dementia`",           "Dementia",                         "Binary",    "Bernoulli(0.04)",
  "Concomitant medications",   "`metformin`",          "Metformin",                        "Binary",    "Bernoulli(0.72)",
  "Concomitant medications",   "`sulfonylurea`",       "Sulfonylurea",                     "Binary",    "Bernoulli(0.35)",
  "Concomitant medications",   "`sglt2i`",             "SGLT2 inhibitor",                  "Binary",    "Bernoulli(0.30)",
  "Concomitant medications",   "`statin`",             "Statin",                           "Binary",    "Bernoulli(0.65)",
  "Concomitant medications",   "`anticoagulant`",      "Anticoagulant",                    "Binary",    "Bernoulli(0.14)",
  "Healthcare utilization",    "`n_hospitalizations`", "No. hospitalizations (prior yr)",  "Count",     "Poisson(0.4)",
  "Healthcare utilization",    "`n_ed_visits`",        "No. ED visits (prior yr)",         "Count",     "Poisson(0.6)",
  "Healthcare utilization",    "`n_distinct_meds`",    "No. distinct medications",         "Count",     "Poisson(8)",
  "Healthcare utilization",    "`comorbidity_index`",  "Combined comorbidity index",       "Continuous","max(Normal(2, 1.5), 0)"
) |>
  gt::gt() |>
  gt::tab_row_group(label = "Healthcare utilization",   rows = Domain == "Healthcare utilization") |>
  gt::tab_row_group(label = "Concomitant medications",  rows = Domain == "Concomitant medications") |>
  gt::tab_row_group(label = "Other comorbidities",      rows = Domain == "Other comorbidities") |>
  gt::tab_row_group(label = "CV / kidney disease",      rows = Domain == "CV / kidney disease") |>
  gt::tab_row_group(label = "Diabetes complications",   rows = Domain == "Diabetes complications") |>
  gt::tab_row_group(label = "Cardiometabolic burden",   rows = Domain == "Cardiometabolic burden") |>
  gt::tab_row_group(label = "Demographics",             rows = Domain == "Demographics") |>
  gt::cols_hide(Domain) |>
  gt_style()
```

## Data analysis

### Context and rationale for analysis plan

The estimand of interest is the **average treatment effect in the overlap (ATO)
population** — patients for whom either treatment would realistically have been
prescribed (clinical equipoise). This is achieved through propensity-score
overlap weighting, which up-weights patients near the decision boundary and
avoids the extrapolation problems of ATT or ATE weighting with extreme scores.

```{r}
#| label: tbl-analysis
#| tbl-cap: "Primary, secondary, and subgroup analysis specification"
#| tbl-subcap:
#|   - "A. Primary analysis"
#|   - "B. Secondary analysis"
#| layout-ncol: 1

# A. Primary
dplyr::tibble(
  Element = c(
    "Hypothesis",
    "Exposure contrast",
    "Outcome",
    "Analytic software",
    "Propensity score model",
    "Confounding adjustment",
    "Outcome model",
    "Standard errors",
    "Missing data",
    "Subgroup analyses"
  ),
  Specification = c(
    "Drug A is not meaningfully different from Drug B on the composite endpoint (non-inferiority).",
    "Drug A vs. Drug B (new users).",
    "Composite: all-cause mortality or HF hospitalization (time-to-event).",
    "R ≥ 4.5; packages: WeightIt, cobalt, gtsummary, survival, ggsurvfit (see Section 9).",
    "Logistic regression: `treatment ~ age + female + race + region + bmi + prior_hf_hosp + diuretic_use + nephropathy + neuropathy + retinopathy + foot_ulcer + insulin_use + pad + afib + cardiomyopathy + ckd + copd + depression + dementia + metformin + sulfonylurea + sglt2i + statin + anticoagulant + n_hospitalizations + n_ed_visits + n_distinct_meds + comorbidity_index`; implemented via `WeightIt::weightit(method = 'glm', estimand = 'ATO')`.",
    "Propensity-score overlap (ATO) weighting; balance assessed with SMDs and Love plot (cobalt).",
    "Weighted Cox PH: `survival::coxph(Surv(time, status) ~ treatment_label, weights = ato_weight, robust = TRUE)`.",
    "Robust (sandwich) standard errors to account for weighting.",
    "No missing data (fully simulated dataset).",
    "None implemented in mock study."
  )
) |>
  gt::gt() |>
  gt_style()

# B. Secondary (not implemented)
dplyr::tibble(
  Element       = c("Note"),
  Specification = c("Secondary analyses are not implemented in this mock study.")
) |>
  gt::gt() |>
  gt_style()
```

```{r}
#| label: tbl-sensitivity
#| tbl-cap: "Sensitivity analyses — rationale, strengths and limitations"

dplyr::tibble(
  Analysis   = c("(None implemented in mock)"),
  Rationale  = c("Sensitivity analyses from the source protocol (e.g. ATT weighting, 90-day follow-up, prior event rate ratio) are not implemented in this mock study."),
  Strengths  = c("N/A"),
  Limitations = c("N/A")
) |>
  gt::gt() |>
  gt_style()
```

## Data sources

### Context and rationale for data sources

**Reason for selection:** Fully simulated data are used to enable reproducible
teaching without the need for restricted real-world data access.

**Strengths:** The simulation is calibrated to approximate a published source
study in terms of cohort size, confounding structure, event rates, and effect
size. It is generated deterministically from a fixed seed (`seed = 20260614`).

**Limitations:** Simulated data cannot capture real-world complexities such as
informative loss to follow-up correlated with covariates, measurement error,
unmeasured confounders, or data source heterogeneity across claims databases.

**Data source provenance:** All data are generated in memory by
`generate_cohort()` in `mock-study/02_data_generation.R`. Nothing is written to
disk.

```{r}
#| label: tbl-data
#| tbl-cap: "Metadata about data sources and software"

dplyr::tibble(
  `Data source`          = c("Simulated cohort"),
  `Study period`         = c("Not applicable (synthetic)"),
  `Eligible entry period`= c("Not applicable"),
  `Data version`         = c("Seed = 20260614; generated 2026-06-14"),
  `Data linkage`         = c("None"),
  `Common data model`    = c("None"),
  `Software`             = c("R ≥ 4.5; renv for package management")
) |>
  gt::gt() |>
  gt_style()
```

## Data management

All data are generated in memory within the R session at analysis time.
No patient data are stored on disk. Package versions are locked via `renv` to
ensure reproducibility across R environments.

## Quality control

The simulated data are validated by inspecting the ATO-weighted results in
`03_primary_endpoint_analysis.qmd`. Target benchmarks: 1-year risk ≈ 3.4%
(Drug B arm), ATO-weighted HR ≈ 0.86 (95% CI includes 1.0).

## Study size and feasibility

The cohort size of n = 28,100 is calibrated to a published source study.
With approximately 500–600 events, the study is
adequately powered to estimate the HR with a 95% CI width of approximately ±0.15
on the log scale.

# Limitation of the methods

1. **Simulated data:** All patient records are synthetic; covariate distributions
   approximate but do not replicate real-world data.
2. **Single primary endpoint only:** Secondary endpoints, subgroup analyses, and
   sensitivity analyses from the source protocol are not implemented.
3. **No negative control outcomes:** Negative control outcome analyses are omitted.
4. **No data source heterogeneity:** The source study pooled three US claims
   databases; this mock uses a single simulated cohort.
5. **Simplified censoring:** A single exponential dropout process is used;
   database-specific disenrollment rules are not modelled.

# Protection of human subjects

This mock study uses only fully synthetic data generated by a random number
generator. No real patient data are used or referenced. No institutional review
board approval is required.

# Reporting of adverse events

Not applicable (non-interventional mock study using simulated data).

# References

::: {#refs}
:::

# Appendices

- **Appendix A:** `mock-study/02_data_generation.R` — defines `generate_cohort()`, the in-memory cohort simulation function.
- **Appendix B:** `mock-study/03_primary_endpoint_analysis.qmd` — primary endpoint analysis report (propensity-score weighting, Table 1, Cox model, cumulative incidence curves).

# Session info

## Loaded packages

```{r}
#| label: session-packages
#| echo: true

subset(
  as.data.frame(utils::installed.packages()[, c("Package", "Version")]),
  Package %in% c("dplyr", "gt", "simsurv", "WeightIt", "cobalt",
                 "survey", "gtsummary", "survival", "ggsurvfit", "ggplot2")
)
```

## Session info

```{r}
#| label: session-info
#| echo: true
utils::sessionInfo()
```

## Repositories

```{r}
#| label: session-repos
#| echo: true
getOption("repos")
```