---
title: "Primary Endpoint Analysis"
subtitle: "Adapted to illustrate agentic code review with Claude Code"
author: "Example Author"
date: today
---

```{r}
#| label: setup
#| include: false

library(dplyr)
library(survival)
library(WeightIt)
library(cobalt)
library(survey)
library(gtsummary)
library(ggsurvfit)
library(ggplot2)
library(scales)
library(knitr)

theme_gtsummary_compact()

# seed for reproducibility of the simulated data (not needed in real-world analyses)
set.seed(20260614)
```

## Background

> **Important:** This is a **mock study** created exclusively to illustrate the
> capabilities of [Claude Code](https://claude.ai/code) for agentic code review
> and validation. Drug names have been replaced with placeholders (**Drug A**,
> **Drug B**). All data are fully simulated. No real-world clinical conclusions
> should be drawn from this analysis.

This document implements a **primary endpoint analysis** following the general
design of a target trial emulation comparing new users of **Drug A** versus
**Drug B** in a chronic disease setting. It uses
entirely synthetic data and generic drug labels.

The **primary outcome** is a composite of **hospitalization for heart failure
(HF) or all-cause mortality**, assessed over up to 52 weeks of follow-up.
Because treatment is not randomized, initiators of the two drugs differ at
baseline (*confounding by indication*). We address this with **propensity-score
overlap (ATO) weighting**, then estimate the treatment effect with a
weighted Cox proportional-hazards model.

```{r}
#| label: load-data
# Generate the simulated cohort in memory by sourcing the data-generation
# function (nothing is read from / written to disk). Resolve the script path
# whether rendered from the project root or from inside mock-study/.
gen_script <- c("02_data_generation.R", file.path("mock-study", "02_data_generation.R"))
source(gen_script[file.exists(gen_script)][1])
cohort <- generate_cohort()

# Confounders to adjust for (baseline covariates).
covariates <- c(
  "age", "female", "race", "region",
  "bmi", "prior_hf_hosp", "diuretic_use",
  "nephropathy", "neuropathy", "retinopathy", "foot_ulcer", "insulin_use",
  "pad", "afib", "cardiomyopathy", "ckd",
  "copd", "depression", "dementia",
  "metformin", "sulfonylurea", "sglt2i", "statin", "anticoagulant",
  "n_hospitalizations", "n_ed_visits", "n_distinct_meds", "comorbidity_index"
)

# Human-readable variable labels (used by gtsummary tables).
var_labels <- list(
  age = "Age, years", female = "Female sex", race = "Race", region = "Region",
  bmi = "Body mass index, kg/m²", prior_hf_hosp = "Prior HF hospitalization",
  diuretic_use = "Diuretic use", nephropathy = "Diabetic nephropathy",
  neuropathy = "Diabetic neuropathy", retinopathy = "Diabetic retinopathy",
  foot_ulcer = "Foot ulcer", insulin_use = "Insulin use",
  pad = "Peripheral arterial disease", afib = "Atrial fibrillation",
  cardiomyopathy = "Cardiomyopathy", ckd = "Chronic kidney disease",
  copd = "COPD", depression = "Depression", dementia = "Dementia",
  metformin = "Metformin", sulfonylurea = "Sulfonylurea", sglt2i = "SGLT2 inhibitor",
  statin = "Statin", anticoagulant = "Anticoagulant",
  n_hospitalizations = "No. hospitalizations (prior yr)",
  n_ed_visits = "No. ED visits (prior yr)",
  n_distinct_meds = "No. distinct medications",
  comorbidity_index = "Combined comorbidity index"
)
```

The analytic cohort comprises **`r format(nrow(cohort), big.mark = ",")`**
patients (`r format(sum(cohort$treatment == 1), big.mark = ",")` Drug A,
`r format(sum(cohort$treatment == 0), big.mark = ",")` Drug B).

## Step 1 — Baseline characteristics (before weighting)

```{r}
#| label: tbl1-unweighted
#| include: false
tbl_unwt <- cohort |>
  select(treatment_label, all_of(covariates)) |>
  tbl_summary(
    by = treatment_label,
    label = var_labels,
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 1
  ) |>
  add_difference(everything() ~ "smd") |>
  modify_header(estimate ~ "**SMD**") |>
  modify_column_hide(columns = conf.low)
```

Several covariates show meaningful imbalance (|SMD| > 0.1), confirming
confounding by indication: Drug A initiators are younger, have higher BMI,
and carry a lower cardiovascular/comorbidity burden.

## Step 2 — Propensity score estimation and overlap (ATO) weighting

We estimate the propensity score (probability of receiving Drug A) with a
logistic regression on all baseline covariates, then derive **overlap (ATO)
weights**. Overlap weights emphasize patients with propensity scores in the
region of clinical equipoise and yield exact mean balance on every covariate
included in the model.

```{r}
#| label: weightit
ps_formula <- reformulate(covariates, response = "treatment")

W <- weightit(
  ps_formula,
  data      = cohort,
  method    = "glm",     # logistic-regression propensity score
  estimand  = "ATO"      # average treatment effect in the overlap population
)

cohort$ato_weight <- W$weights
W
```

### Covariate balance before vs. after weighting

```{r}
#| label: love-plot
#| fig-width: 8
#| fig-height: 8
love.plot(
  W,
  stats        = "mean.diffs",
  abs          = TRUE,
  thresholds   = c(m = 0.1),
  var.order    = "unadjusted",
  drop.distance = TRUE,
  sample.names = c("Unweighted", "ATO weighted"),
  title        = "Covariate balance: standardized mean differences"
)
```

After overlap weighting, all standardized mean differences are driven to ~0,
indicating excellent balance.

### Propensity score overlap

```{r}
#| label: ps-overlap
#| fig-width: 8
#| fig-height: 4.5
bal.plot(
  W,
  var.name = "prop.score",
  which    = "both",
  type     = "density",
  sample.names = c("Unweighted", "ATO weighted")
) +
  labs(title = "Propensity score distribution by treatment group",
       x = "Propensity score (P[Drug A])")
```

### Baseline characteristics — before and after ATO weighting

```{r}
#| label: tbl1-weighted
#| include: false
svy <- svydesign(ids = ~1, weights = ~ato_weight, data = cohort)

tbl_wt <- svy |>
  tbl_svysummary(
    by      = treatment_label,
    include = all_of(covariates),
    label   = var_labels,
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 1
  ) |>
  add_difference(everything() ~ "smd") |>
  modify_header(estimate ~ "**SMD**") |>
  modify_column_hide(columns = conf.low)
```

```{r}
#| label: tbl1-merged
tbl_merge(
  tbls        = list(tbl_unwt, tbl_wt),
  tab_spanner = c("**Unweighted**", "**ATO weighted**")
) |>
  modify_caption(
    "**Table 1. Baseline characteristics before and after ATO overlap weighting**"
  )
```

## Step 3 — Primary endpoint analysis (weighted Cox model)

The treatment effect on the composite endpoint is estimated with a Cox
proportional-hazards model incorporating the overlap weights. Robust
(sandwich) standard errors account for the weighting.

```{r}
#| label: cox-model
cox_wt <- coxph(
  Surv(time, status) ~ treatment_label,
  data    = cohort,
  weights = cohort$ato_weight,
  robust  = TRUE
)

tbl_regression(cox_wt, exponentiate = TRUE) |>
  modify_header(estimate ~ "**HR**") |>
  modify_caption(
    "**Table 2. Primary composite endpoint — ATO-weighted hazard ratio**"
  )
```

```{r}
#| label: hr-extract
hr      <- exp(coef(cox_wt))[["treatment_labelDrug A"]]
ci      <- exp(confint(cox_wt))["treatment_labelDrug A", ]
hr_text <- sprintf("%.2f (95%% CI, %.2f-%.2f)", hr, ci[1], ci[2])
```

**Primary result:** Drug A vs. Drug B HR = **`r hr_text`** for the
composite of HF hospitalization or all-cause mortality.

## Step 4 — One-year cumulative incidence

```{r}
#| label: one-year-risk
fit_wt <- survfit(
  Surv(time, status) ~ treatment_label,
  data    = cohort,
  weights = cohort$ato_weight
)

risk_1yr <- summary(fit_wt, times = 365)
risk_tbl <- data.frame(
  Group         = sub("treatment_label=", "", risk_1yr$strata),
  `1-year risk` = sprintf("%.1f%% (95%% CI, %.1f%%-%.1f%%)",
                          100 * (1 - risk_1yr$surv),
                          100 * (1 - risk_1yr$upper),
                          100 * (1 - risk_1yr$lower)),
  check.names = FALSE
)
kable(risk_tbl, caption = "Table 3. ATO-weighted 1-year risk by arm")
```

```{r}
#| label: cif-plot
#| fig-width: 8
#| fig-height: 5
survfit2(
  Surv(time, status) ~ treatment_label,
  data    = cohort,
  weights = cohort$ato_weight
) |>
  ggsurvfit(type = "risk", linewidth = 1) +
  add_confidence_interval() +
  add_risktable(risktable_stats = "n.risk") +
  coord_cartesian(xlim = c(0, 365)) +
  scale_x_continuous(breaks = seq(0, 365, 90)) +
  scale_y_continuous(labels = percent) +
  labs(
    title = "Cumulative incidence of HF hospitalization or all-cause mortality",
    x = "Days since treatment initiation",
    y = "Cumulative incidence"
  )
```

### Interpretation

After propensity-score overlap weighting, the risk of
the composite endpoint is **not meaningfully different** between Drug A and
Drug B initiators, with a hazard ratio close to the published estimate of
0.86 and a confidence interval that crosses 1.0. The crude (unweighted) analysis
would have overstated a Drug A benefit because of confounding by
indication — illustrating why balancing the treatment groups is essential.
