---
title: "Validation Report"
subtitle: "Unit-test results for the mock-study analytical pipeline"
author: "Example Author"
date: today
---

```{r}
#| label: setup
#| include: false

library(testthat)
library(dplyr)
library(gt)
library(here)

# Suppress CLI colour codes in rendered output
options(cli.num_colors = 0L)
```

> **Important:** This is a **mock validation report** for the synthetic teaching
> study described in `01_csp.qmd`. It runs the unit tests in `test/` and
> renders results inline. No real patient data are involved.

This document executes the three `testthat` test suites that validate the
mock-study analytical pipeline and renders their results as formatted tables.
Tests are organised as follows:

| Test file | Scope |
|---|---|
| `test_data_generation.R` | Internal correctness of `generate_cohort()` |
| `test_primary_endpoint_analysis.R` | Correctness of the PS-weighting and Cox-model pipeline |
| `test_protocol_adherence.R` | Adherence to pre-specified values in `01_csp.qmd` |

```{r}
#| label: helpers
#| include: false

# Run one test file silently and return its data.frame result.
run_tests <- function(file) {
  testthat::test_file(
    here("test", file),
    reporter = "silent",
    stop_on_failure = FALSE
  ) |>
    as.data.frame() |>
    dplyr::select(test, nb, passed, failed, skipped, error, real) |>
    dplyr::rename(
      `Test`         = test,
      `Expectations` = nb,
      `Passed`       = passed,
      `Failed`       = failed,
      `Skipped`      = skipped,
      `Error`        = error,
      `Time (s)`     = real
    ) |>
    dplyr::mutate(
      Status   = dplyr::case_when(
        Error           ~ "✖ Error",
        Failed > 0      ~ "✖ Fail",
        Skipped == Expectations ~ "○ Skip",
        TRUE            ~ "✔ Pass"
      ),
      `Time (s)` = round(`Time (s)`, 3)
    )
}

# Render a result data.frame as a gt table.
result_table <- function(df, title) {
  n_pass <- sum(df$Status == "✔ Pass")
  n_fail <- sum(df$Failed > 0 | df$Error)
  n_skip <- sum(df$Skipped > 0 & df$Failed == 0 & !df$Error)

  df |>
    dplyr::select(Status, Test, Expectations, Passed, Failed, `Time (s)`) |>
    gt() |>
    tab_header(
      title    = title,
      subtitle = glue::glue("{n_pass} passed · {n_fail} failed · {n_skip} skipped")
    ) |>
    cols_align(align = "left",   columns = c(Status, Test)) |>
    cols_align(align = "center", columns = c(Expectations, Passed, Failed, `Time (s)`)) |>
    tab_style(
      style     = cell_text(color = "#2e7d32", weight = "bold"),
      locations = cells_body(columns = Status,
                             rows    = Status == "✔ Pass")
    ) |>
    tab_style(
      style     = cell_text(color = "#c62828", weight = "bold"),
      locations = cells_body(columns = Status,
                             rows    = grepl("✖", Status))
    ) |>
    tab_style(
      style     = cell_fill(color = "#fff3e0"),
      locations = cells_body(rows = grepl("✖", Status))
    ) |>
    tab_options(table.width = pct(100), data_row.padding = px(3))
}
```

---

## 1 — Data generation (`test_data_generation.R`)

Tests internal correctness of `generate_cohort()`: output structure, column
types, value ranges, marginal plausibility, and reproducibility.

```{r}
#| label: run-data-generation
res_data <- run_tests("test_data_generation.R")
result_table(res_data, "generate_cohort() unit tests")
```

---

## 2 — Primary endpoint analysis (`test_primary_endpoint_analysis.R`)

Tests the PS-weighting and Cox-model pipeline: weight validity, covariate
balance, Cox coefficient extraction, and 1-year risk formatting.

```{r}
#| label: run-primary-analysis
res_primary <- run_tests("test_primary_endpoint_analysis.R")
result_table(res_primary, "Primary endpoint analysis unit tests")
```

---

## 3 — Protocol adherence (`test_protocol_adherence.R`)

Verifies that every pre-specified value in `01_csp.qmd` is honoured by the
implementation: default parameters, covariate distributions, analysis choices,
and quality-control benchmarks (n = 28 100; may take ~20 s).

```{r}
#| label: run-protocol-adherence
res_protocol <- run_tests("test_protocol_adherence.R")
result_table(res_protocol, "Protocol-adherence unit tests")
```

---

## Overall summary

```{r}
#| label: overall-summary

file_labels <- c(
  "test_data_generation.R"         = "Data generation",
  "test_primary_endpoint_analysis.R" = "Primary endpoint analysis",
  "test_protocol_adherence.R"      = "Protocol adherence"
)

list(
  "test_data_generation.R"           = res_data,
  "test_primary_endpoint_analysis.R" = res_primary,
  "test_protocol_adherence.R"        = res_protocol
) |>
  purrr::imap(function(df, file) {
    dplyr::summarise(
      df,
      Suite        = file_labels[[file]],
      Tests        = dplyr::n(),
      Expectations = sum(Expectations),
      Passed       = sum(Passed),
      Failed       = sum(Failed),
      Skipped      = sum(Skipped),
      `Time (s)`   = round(sum(`Time (s)`), 1),
      Status       = dplyr::if_else(
        sum(Failed) + sum(Error) == 0,
        "✔ All pass",
        glue::glue("✖ {sum(Failed)} failed")
      )
    )
  }) |>
  purrr::list_rbind() |>
  gt() |>
  tab_header(title = "Validation summary") |>
  cols_align(align = "left",   columns = c(Suite, Status)) |>
  cols_align(align = "center", columns = c(Tests, Expectations, Passed, Failed, Skipped, `Time (s)`)) |>
  tab_style(
    style     = cell_text(color = "#2e7d32", weight = "bold"),
    locations = cells_body(columns = Status, rows = grepl("✔", Status))
  ) |>
  tab_style(
    style     = cell_text(color = "#c62828", weight = "bold"),
    locations = cells_body(columns = Status, rows = grepl("✖", Status))
  ) |>
  grand_summary_rows(
    columns  = c(Tests, Expectations, Passed, Failed, Skipped),
    fns      = list(Total = ~sum(.)),
    fmt      = ~ fmt_integer(.)
  ) |>
  tab_options(table.width = pct(100), data_row.padding = px(3))
```

---

## Session info

```{r}
#| label: session-info
utils::sessionInfo()
```
